Seniority and Hierarchy Configuration Interaction for Radicals and \\ Excited States

### 2023-10-24 Manuscript/Response_Letter/Response_Letter.tex

{6) The authors keep mentioning ``reference theoretical values.'' Please mention what kind of reference values were used.
}

\alert{
This information is provided in the Supporting Information, which is stated on page x of the revised manuscript, reading
``For each excited state considered here, the reference value and the corresponding method to obtain it (of high-order coupled cluster or extrapolated FCI quality) can be found in the Supporting Information.''
}

{7) The authors look at several small-sized molecules for which FCI can still be determined. There is nothing wrong with that. However, I am missing numerical studies on slightly larger systems, which are known to be challenging in computational chemistry. I recommend adding some ``larger'' molecules, which feature open- and closed-shell ground and excited states. Examples are cyclobutadiene (its automatization problem) and o-, m-, and p-benzynes. These systems are notoriously difficult to study and extensively discussed in the literature. Including just those four molecules would add more value to the paper and allow a direct comparison to other state-of-the-art computational methods.
}

\alert{
This is a pertinent remark.
From the examples suggested by the reviewer, we decided to investigate the automerization problem of cyclobutadiene.
Our results can be found in a new paragraph on page x, starting as:
``We further explored hCI to describe the automerization barrier in the ground state of cyclobutadiene...''
On the other hand, the case of benzynes is left as a perspective for the future, otherwise the manuscript would be excessively lengthy in our opinion.
We notice that extensive CI+EN2 additional calculations have been carried out and a corresponding new subsection (IV.E) has been added to the manuscript, based on a suggestion from Reviewer \#1.
We further highlight the point raised by Reviewer \#2 in the Conclusion, reading
``hCI also produced reasonable energies for the automerization barrier of cyclobutadiene.
It remains to be seen how these models perform for other strongly correlated systems, larger molecules, and charge transfer excited states.''
}

### 2023-10-24 Manuscript/manuscript.tex
cyclobutadiene

@@

We further gauged the performance of hCI to describe excited states of closed- and open-shell systems, based on either HF ground-state orbitals or state-specific orbitals.
For excitations of closed-shell systems, $\Delta$hCI2 and $\Delta$hCI3 are comparable to their excitation-based counterparts, $\Delta$CISD and $\Delta$CISDT,
whereas $\Delta$hCI1, $\Delta$hCI1.5, and $\Delta$hCI2.5 are inaccurate bearing in mind their computational cost.
The poor performance at lower orders is ascribed to the different minimal references employed,
a single seniority-zero determinant for the ground state and a single seniority-two CSF for the excited state,
which introduces a strong bias on the classes of determinants accessed in the hCI calculations.
$\Delta$hCI performs significantly better for the doublet transitions when compared to the results for closed shells.
In this case, the reference of both ground and excited states comprises the same type of single seniority-one determinant.
The advantage of using state-specific orbitals over ground-state ones depends on the choice of reference and the order in which the CI is truncated.
When similar references are adopted (as for the doublet transitions), such advantage is already evident at $\Delta$hCI1.
In contrast, for unlike references (as for the closed-shell excitations), the state-specific approach only becomes advantageous at somewhat higher orders ($\Delta$hCI2).
The present findings highlight the challenge in describing, on an equal footing, states with qualitatively different characters, in particular ground and singly-excited states of closed-shell systems.
In this regard, rotations within occupied and virtual orbitals represent yet another factor worth exploring in the future developments of hCI models.

>>>>

\fk{
We further explored hCI to describe the automerization barrier in the ground state of cyclobutadiene,
a well-known and challenging problem that involves strong correlation. \cite{Monino_2022}
High-level calculations are needed to achieve quantitative values for the height of the barrier that 
connects the two equivalent rectangular $D_{2h}$ equilibrium geometries through the square $D_{4h}$ transition state geometry. \cite{Monino_2022}
Here we employ the geometries presented in Ref.~\onlinecite{Monino_2022}, the 6-31+G(d) basis set and the frozen-core approximation.
We optimize the orbitals for two closed-shell determinants in the $D_{2h}$ geometry, and for two spin-flipped open-shell determinants in the $D_{4h}$ geometry, from which the excited determinants were generated in the CI calculations.
For this basis set, the reference value of \SI{7.51}{kcal/mol} was obtained at CC with single, double, triple, and quadruple excitations. \cite{Monino_2022}
With respect to this reference value, the errors for the automerization barrier are
\SI{+3.55}{kcal/mol} with our two-determinant calculations, \SI{-2.28}{kcal/mol} with hCI2, \SI{-1.07}{kcal/mol} with hCI2+EN2, \SI{+0.61}{kcal/mol} with CISD, and \SI{-0.49}{kcal/mol} with CISD+EN2.
These values are comparable to those obtained with other methods, extensively discussed in Ref.~\onlinecite{Monino_2022}.
Here we just mention the errors of CC with singles and doubles, \SI{+0.80}{kcal/mol}, and of spin-flip equation-of-motion CC (EOM-CC) with singles and doubles, \SI{-1.65}{kcal/mol},
which share the same $\order*{N^6}$ computational scaling as CISD and hCI2.
This specific example serves to illustrate that the accuracy of hCI models can be similar to that of other methods.
As such, hCI can be taken into consideration in future systematic studies that tackle challenging chemical problems where strong correlation comes into play.
}

@@

We further gauged the performance of hCI to describe excited states of closed- and open-shell systems, based on either HF ground-state orbitals or state-specific orbitals.
For excitations of closed-shell systems, $\Delta$hCI2 and $\Delta$hCI3 are comparable to their excitation-based counterparts, $\Delta$CISD and $\Delta$CISDT,
whereas $\Delta$hCI1, $\Delta$hCI1.5, and $\Delta$hCI2.5 are inaccurate bearing in mind their computational cost.
The poor performance at lower orders is ascribed to the different minimal references employed,
a single seniority-zero determinant for the ground state and a single seniority-two CSF for the excited state,
which introduces a strong bias on the classes of determinants accessed in the hCI calculations.
$\Delta$hCI performs significantly better for the doublet transitions when compared to the results for closed shells.
In this case, the reference of both ground and excited states comprises the same type of single seniority-one determinant.
The advantage of using state-specific orbitals over ground-state ones depends on the choice of reference and the order in which the CI is truncated.
When similar references are adopted (as for the doublet transitions), such advantage is already evident at $\Delta$hCI1.
In contrast, for unlike references (as for the closed-shell excitations), the state-specific approach only becomes advantageous at somewhat higher orders ($\Delta$hCI2).
The present findings highlight the challenge in describing, on an equal footing, states with qualitatively different characters, in particular ground and singly-excited states of closed-shell systems.
In this regard, rotations within occupied and virtual orbitals represent yet another factor worth exploring in the future developments of hCI models.
\fk{hCI also produced reasonable energies for the automerization barrier of cyclobutadiene.
It remains to be seen how these models perform for other strongly correlated systems, larger molecules, and charge transfer excited states.}
